# Preface {-}

## Training overview

In this course, we will present concepts, theory, software and a Shiny interface.
Mainly we will focus on designs that you might consider for time-to-event endpoints.
In addition to *classical* approaches assuming a proportional hazards assumption, we will provide methods for designing under non-proportional hazards assumptions.
While most studies still use a logrank test, we will also touch on some alternatives along with their potential advantages and disadvantages.

::: {.rmdnote}
**Disclaimer**

All opinions expressed are those of the presenters and not Merck & Co., Inc., Kenilworth, NJ, USA.
:::

## Chapters and training sections {-}

- Background theory (30 minutes)
  - Extension to non-proportional hazards
  - Group sequential design asymptotic distribution
  - Spending function bounds

- Proportional hazards applications with Shiny app (40 minutes)
  - Lachin and Foulkes method for sample size derivation
  - Design setup with exponential distribution
  - Design setup with cure model
  - Updating bounds at time of analysis
  - Event-based and calendar-based spending bounds
  - Exercise

- Break (15 minutes)

- Non-proportional hazards model with logrank test (60 minutes)
  - Piecewise model
  - Average hazard ratio
  - Statistical information and time
  - Introduction to gsdmvn, gsDesign2 and simtrial

- Break (10 minutes)

- Weighted logrank, RMST and combination tests (55 minutes)
  - Introduction to methods
  - Weighted logrank
  - MaxCombo
  - Exercise

## Software and supporting materials

- Useful directories in course repository at *LINK*:
  - `data/`: contains design files for examples; also simulation results
  - `vignettes/`: reports produced by Shiny app to summarize designs
  - `simulation/`: R code and simulation data for the last part of the course

## Installing R packages

If you choose to install R packages locally:

- The **gsDesign** package (v3.2.1) is available at CRAN

```{r, eval=FALSE}
install.packages("gsDesign")
```

- For non-proportional hazards, the following 3 packages would be useful to install

```{r, eval=FALSE}
remotes::install_github("Merck/gsdmvn")
remotes::install_github("Merck/gsDesign2")
remotes::install_github("Merck/simtrial")
```

You will need reasonably recent versions of R and packages.
