# Other tests

## Outline

We consider other alternative tests for group sequential design.

- Review fixed design
  - Weighted logrank test
  - MaxCombo test
  - Illustration using [`npsurvSS`](https://cran.r-project.org/package=npsurvSS)

- Group sequential design with weighted logrank test
  - Under a given boundary
  - Boundary calculation
  - Illustration using [`gsdmvn`](https://merck.github.io/gsdmvn/)

- Group sequential design with MaxCombo test
  - Under a given boundary
  - Boundary calculation
  - Illustration using [`gsdmvn`](https://merck.github.io/gsdmvn/)

## Fixed design

For simplicity, we made a few key assumptions.

- Balanced design (1:1 randomization ratio).
- 1-sided test.
- Local alternative: variance under null and alternative are approximately equal.
- Accrual distribution: Piecewise uniform.
- Survival distribution: piecewise exponential.
- Loss to follow-up: exponential.
- No stratification.
- No cure fraction.

The fixed design part largely follows the concept described in @yung2019sample.

## Notation

- $\alpha$: Type I error
- $\beta$: Type II error or power (1 - $\beta$)
- $z_\alpha$: upper $\alpha$ percentile of standard normal distribution
- $z_\beta$: upper $\beta$ percentile of standard normal distribution

We considered a 1-sided test with type I error at $\alpha=0.025$ and $1-\beta=80\%$ power.

```{r}
z_alpha <- abs(qnorm(0.025))
z_alpha
```

```{r}
z_beta <- abs(qnorm(0.2))
z_beta
```

## Sample size calculation

- $\theta$: effect size
- $n$: total sample size
- $Z$: test statistics is asymptotic normal
  - Under null hypothesis: $Z \sim \mathcal{N}(0, \sigma_0^2)$
  - Under alternative hypothesis: $Z \sim \mathcal{N}(\sqrt{n}\theta, \sigma_1^2)$

By assuming local alternative, we have

$$\sigma_0^2 \approx \sigma_1^2 = \sigma^2$$
In this simplified case, the sample size can be calculated as

$$ n = \frac{4 (z_{\alpha}+z_{\beta})^{2}}{\theta^2} $$

## Examples

- [Two-sample t-test](https://keaven.github.io/gsd-deming/fixed-design.html#two-sample-t-test)
- [Logrank test](https://keaven.github.io/gsd-deming/fixed-design.html#logrank-test)

## Sample size calculation under non-proportional hazards

- <https://keaven.github.io/gsd-deming/fixed-design.html#non-proportional-hazards>

## Fixed design with weighted logrank test

- [Examples](https://keaven.github.io/gsd-deming/fixed-design.html#weighted-logrank-test)
- [Technical details](https://keaven.github.io/gsd-deming/fixed-design.html#technical-details) are skipped

## Fixed design with MaxCombo test

- [Examples](https://keaven.github.io/gsd-deming/fixed-design.html#maxcombo-test)


## Weighted logrank test

Similar to the fixed design, we can define the test statistics for weighted logrank test using counting process formula

$$ Z_k=\sqrt{\frac{n_{0}+n_{1}}{n_{0}n_{1}}}\int_{0}^{t_k}w(t)\frac{\overline{Y}_{0}(t)\overline{Y}_{1}(t)}{\overline{Y}_{0}(t)+\overline{Y}_{0}(t)}\left\{ \frac{d\overline{N}_{1}(t)}{\overline{Y}_{1}(t)}-\frac{d\overline{N}_{0}(t)}{\overline{Y}_{0}(t)}\right\} $$

Note, the only difference is that the test statistics fixed analysis up to $t_k$ at $k$-th interim analysis

## Group sequential design with weighted logrank test

- [Examples with pre-defined boundary](https://keaven.github.io/gsd-deming/wlr.html#example-scenario)
- [Examples with calculated boundary](https://keaven.github.io/gsd-deming/wlr-boundary.html)
- [Technical details](https://keaven.github.io/gsd-deming/wlr.html#technical-details-1) are skipped

## Group sequential design with MaxCombo test

- [Examples with pre-defined boundary](https://keaven.github.io/gsd-deming/wlr.html#example-scenario)
- [Examples with calculated boundary](https://keaven.github.io/gsd-deming/maxcombo-boundary.html)
- [Technical details](https://keaven.github.io/gsd-deming/maxcombo.html#outline-of-technical-details) are skipped
