---
title: EXAMINE non-inferiority design
subtitle: Includes bound update at interim analysis
output: html_document
---


# Design assumptions

The EXAMINE trial provides an example of how to design a group sequential trial for a non-infereriority trial.
The key difference for the design is that we are not trying to rule out an excessive event rate in the experimental treatment group relative to control (HR $\ge 1.3$) when in truth there is no difference in the time-to-event distribution for the control and experimental groups (HR$=1$).
Here we have a one-sided design. 
The primary endpoint is evaluated using repeated confidence intervals as noted in study publications.
In general, other parameters are also from study publications.
Documentation of design parameters is more complete in the course book.


It may be useful to add comments to the following to document what all of the parameters are.
We can only provide a couple of comments here that are particularly relevant to a non-inferiority design.

```{r}
library(gsDesign)
library(dplyr)
library(gt)
k <- 3
test.type <- 1
alpha <- 0.025
beta <- 0.09341
astar <- 0
timing <- c(0.8461538, 0.9230769)
sfu <- sfLDOF
sfupar <- c(0)
sfl <- sfLDOF
sflpar <- c(0)
# The following 2 parameters are notable for a non-inferiority design
hr <- 1 # Alternate hypothesis is equal time-to-event distributions
hr0 <- 1.3 # Null hypothesis we are trying to reject is HR >= 1.3
eta <- 0.01005034
gamma <- c(10)
R <- c(2)
S <- NULL
T <- 4.75
minfup <- 2.75
ratio <- 1
obs <- matrix(c(100, 0.03562718), ncol = 2)
obs <- obs[(!is.na(obs[, 1])) & (!is.na(obs[, 2])), 2]
lambdaC <- obs
```

Alternatively, you can access any parameter from above an document here what it is.
For instance, the assumed exponential failure rate for the control group is `r lambdaC` and the trial is powered at `r 100*(1-beta)`% to rule out a hazard ratio of `r hr0` (Experimental/Control; null hypothesis) under the assumption the true hazard is `r hr` (alternate hypothesis) with 1-sided Type I error `r alpha`.




# Derive group sequential design

With the above parameters, we derive a group sequential design.

```{r}
x <- gsSurv(
  k = k,
  test.type = test.type,
  alpha = alpha,
  beta = beta,
  astar = astar,
  timing = timing,
  sfu = sfu,
  sfupar = sfupar,
  sfl = sfl,
  sflpar = sflpar,
  lambdaC = lambdaC,
  hr = hr,
  hr0 = hr0,
  eta = eta,
  gamma = gamma,
  R = R,
  S = S,
  T = T,
  minfup = minfup,
  ratio = ratio
) 
x %>% 
gsBoundSummary(ratio = 1, digits = 4, ddigits = 2, tdigits = 1, timename = 'Month') %>%
  gt() %>%
  tab_header(title = "EXAMINE trial original design approximation")
```
We note that some of the parameters where chosen so that the design event counts at analyses are just below integer values:

```{r}
x$n.I
```

# Update bounds at time of analysis

We enter the events at the interim analysis for the trial.
Since `gsDesign()` requires at least 2 analyses, we also provide the final analysis planned event count even though that was not needed at the time of the interim analysis.

```{r}
n.I <- c(621, 650)
ku <- length(n.I)
# This is just specifying event fraction vs final planned
# is used for computing spending.
# If calendar spending were specified, calendar fraction 
# would be used here.
usTime <- n.I / x$n.I[x$k]
lsTime <- usTime
```

Now we incorporate this to update the original design.

```{r}
xu <- gsDesign(
  k = ku,
  test.type = test.type,
  alpha = x$alpha,
  beta = x$beta,
  astar = astar,
  timing = timing,
  sfu = sfu,
  sfupar = sfupar,
  sfl = sfl,
  sflpar = sflpar,
  n.I = n.I,
  maxn.IPlan = x$n.I[x$k],
  delta = x$delta,
  delta1 = x$delta1,
  delta0 = x$delta0,
  usTime = usTime,
  lsTime = lsTime
)
```

Next we document the updated bounds.
The first code line here is needed for any case where `hr0 != 1`; i.e. for non-inferiority trials or, in the case of vaccines, super-superiority trials.

```{r}
# The first line is required to make gsBoundSummary to work correctly
# for non-inferiority and super-superiority trials
xu$hr0 <- x$hr0
gsBoundSummary(
  xu,
  deltaname = "HR",
  logdelta = TRUE,
  Nname = "Events",
  digits = 4,
  ddigits = 2,
  tdigits = 1,
  exclude = c(
    "B-value", "CP", "CP H1", "PP")) %>%
  gt() %>%
  tab_header(
    title = "EXAMINE trial design",
    subtitle = "Update at time of analysis"
  )
```

# Environment information

- Run date: `r Sys.time()`
- R version: `r R.version.string`
- gsDesign version: `r as.character(packageVersion("gsDesign"))`
